When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-17 doc/paper/abstract.tex
spell check done and makefile updated
now proofread to send off to Pierre and Changjie

Increasing vapor pressure deficit (VPD) increases atmospheric demand for
water, and vapor pressure deficit is expected to rise with increasing
greenhouse gases. While increased evapotranspiration (ET) in response to
increased atmospheric demand seems intuitive, plants are capable of
reducing ET in response to increased VPD by closing their stomata, in an
effort to conserve water. Here we examine which effect dominates
response to increasing VPD: atmospheric demand and increases in ET, or
plant physiological response (stomata closure) and decreases in ET. We
use Penman-Monteith, combined with semi-empirical optimal stomatal
regulation theory and underlying water use efficiency, to develop a
theoretical framework for understanding how ET responds to increases in
VPD.

The theory suggests that for most environmental conditions and plant
types, plant physiological response dominates and ET decreases with
increasing VPD. Plants that are evolved or bred to prioritize primary
production over water conservation (e.g. crops) exhibit a higher
likelihood of atmospheric demand-driven response (ET
increasing). However for forest, grass, savannah, and shrub plant types,
ET more frequently decreases than increases with rising VPD. This work
serves as an example of the utility of our simplified framework for
disentangling land-atmosphere feedbacks, including
the characterization of ET response in an atmospherically drier,
enriched CO$_2$ world.

>>>>

Increasing vapor pressure deficit (VPD) increases atmospheric demand
for water, and vapor pressure deficit is expected to rise with
increasing greenhouse gases. While increased evapotranspiration (ET)
in response to increased atmospheric demand seems intuitive, plants
are capable of reducing ET in response to increased VPD by closing
their stomata, in an effort to conserve water. Here we examine which
effect dominates response to increasing VPD: atmospheric demand and
increases in ET, or plant physiological response (stomata closure) and
decreases in ET. We use Penman-Monteith, combined with semi-empirical
optimal stomatal regulation theory and underlying water use
efficiency, to develop a theoretical framework for understanding how
ET responds to increases in VPD. The theoretical ET response to VPD
over a range of reasonable environmental conditions and plant
characteristics varies from a strong decrease in ET in response to
increasing VPD (water conservative) to a strong increase in ET in
response to VPD (water intensive), highlighting the diversity of plant
water regulation strategies.

The ET response varies due to: 1) climate, with tropical and
temperate climates more likely to exhibit a positive ET response than
boreal and arctic climates; 2) photosynthesis strategy, with C3 plants
more likely to exhibit a positive ET response than C4 plants; and 3)
due to plant type, with crops more likely to exhibit a positive ET
response, and shrubs and gymniosperm trees more likely to exhibit a
negative ET response. These results, derived from previous literature
connecting plant parameters to plant and climate characteristics,
highlight the utility of our simplified framework for understanding
complex land atmosphere systems in terms of idealized scenarios.

### 2019-02-17 doc/paper/body.tex
spell check done and makefile updated
now proofread to send off to Pierre and Changjie

  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET reponse to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

In the ``sign term'' most of quantities are relatively constant,
except for VPD and the plant parameters g$_1$ and uWUE. In the scaling
term, most of the terms are relatively constant with the exception of
aerodynamic conductance ($g_a$) and temperature (especially its effect
through $\Delta$). To determine the range of probable ET reponses to
VPD we will systematically vary this parameters according to Table
\ref{param_varying}, while all other parametrs are held fixed (Table
\ref{param_fixed}). Physical variables ($g_a$, $T$) and plant
physiological paramters (g$_1$, uWUE) are varied according to
literature-based expectations for a range of growing season conditions
and plant types \citep{Zhou_2015, Medlyn_2017}. Using this previous
literature we can connect the effect of varying plant parameters to
specific plant types and characteristics.

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided in cases where the quantites are directly derived from
    previous literature. Conceptual percentiles
    for interpreting each quantity are noted.}
  \label{param_varying}
  \centering
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min (~15th percentile) & med (~50th percentile) & max
                                                                                            (~15th
                                                                                            percentile)
    & citation  \\
    \hline
    \input{param_varying.tex}
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}
  \centering \includegraphics{./fully_idealized)sign.pdf}
  \caption{The functional form of the $\frac{\partial \; ET}{\partial
      \; VPD}$ evaluated at the median value of the scaling term, for
    varying values of g$_1$ and uWUE as given in Table \ref{param_varying}.}
  \label{sign}
\end{figure}

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (through $\Delta$ controls the efficiency of
energy conversion to latent heat), amplify or supress the plant
physiological response represented in the sign term. Both lower
temperatures and higher aerodynamic conductance lead to amplified ET
response to VPD, with the variabilty of aerodynamic conudcance
resulting in a slightly higher variability of ET response relative to
temperature, which is still an important factor.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
dervatives for understanding the behavior of complex systems. Our
results are a conceptual tool for answering the question, ``What is
the ET response to VPD with all other quantities held fixed?'' This is
useful, and we would argue is critical, for understanding complex
responses when many quantities are varying symotaneously. Through
partial derivatives, we can approach answering questions like ``What
is the relative contribution of VPD in some observed response given
that X Y and Z were also varying?'' In this example, X Y and Z could
be, for example, radiation, soil moisture, and temperature. Equation
\ref{d_et} explicitly provides an estimate of the ET response to VPD,
given assumptions about ecosystem plant characteristics through the
parameters g$_1$ and uWUE. Equation \ref{d_et} is generic to external
environmental factors and the timescale over which a VPD response
acts, and has a straightforward conceptual meaning. It provides the ET
response, given other quantities and parameters held fixed. So, while
the plant parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and that they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}) it is
useful to rexamine how these quantities vary in previous literature,
framed by our results on ET responses. In \citet{Zhou_2015} most plant
types have uWUE similiar to our median value, where whether ET
response is postive (water intensive) or negative (water conservative)
depends on VPD. However, crops, which we expect to prioritize carbon
uptake over water conservation, have a higher uWUE closer to our
maximum value. This higher uWUE results in a higher liklhood for an
increasing ET response to VPD, which intuitively amkes sense given
taht we expect crops ro keep stoamta open for access to carbon, at the
cost of water loss. Shrubs, and to a lesser extent grass, gave an uWUE
closer to the minumum uWUE. For these plant types, we would then
expect a higher liklhood of a water conservatve response; that is, a
decrease of ET in response to VPD. It is important to note that, while
\citet{Zhou_2015} did not examine the roll of soil moisture in within
plant-type variability of uWUE, we might expect some variation,
especially in extreme cases when soil water becomes a limiting factor.

In \citet{Medlyn_2017} there is not as robust of a relationship
between plant type and g$_1$ as there was for uWUE in
\citet{Zhou_2015}. Given that there is a robust relatiohsip between
g$_1$ and plant type in \citet{Lin_2015}, the ambiguos results in
\citet{Medlyn_2017} could be due to noise in ecosystem scale
observations, the consequences of imposing a model structure in some
of the g$_1$ observations, or real ecosystem-scale within-plant type
variability. These differences between leaf-scale behavior and
ecosystem scale behaviro highlight the importance of understanding how
and why slope parameters vary at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can make some
connections between g$_1$-induced ET response variability and plant
characteristics, which are subject to the strong caveat that we still
do not fully understand the g$_1$'s utility at the ecosystem
scale. Extrapolating \citet{Lin_2015}'s results (e.g. Fig. 2) to the
ecosystem scale would reveal the following relationships between g$_1$
and plant types:

\begin{itemize}
  \item C3 plants would have a generally higher g$_1$ than C4 plants,
  \item Crops would have a larger g$_1$ than shrubs, grass, and
    angiosperm trees, which have a higher g$_1$ than gymnosperm
    trees.
  \item Tropical and temperate climates would generally host plants
    with a higher g$_1$ than artic and boreal climates.
\end{itemize}
For all of the above relationships, a higher g$_1$ means a higher
liklihood of a postive ET response to increasing VPD (water
intensive), and a smaller VPD dependence of the ET response.

Our theoretical results, when framed by previous literature tying
parametets to plant types and climate, highlight the variability of ET
response to VPD as a function of plant characteristics. Whether an
ecosystem increases or decreases ET in response to VPD will vary
depending on vegetation and climate. Generalizations about ET response
to VPD therefore require thoughtful consideration of ecosystem
characteristics; statements such as ``ET increases with warming due to
increases in VPD'' may be false depending on the water conservation
strategy of a given ecosystem **CITE SOME EXAMPLES OF THIS LOGIC FROM
GLOBAL WARMING LIT**.

Additionally, the sensitivity of ET reponse to plant parameter
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf and ecosystem scale
are not directly analgalous much be understood. While the derivation
here was explicit about the assumptions of a constant g$_1$ term with
respect to the VPD perturbation, many sophsiticated land surface
models and earth system models employing stomatal conductance models
make similiar, and even stronger assumptions about the stationarity of
VPD slope terms. In particular, many models will assume a constant
g$_1$ term within a given plant functional type, and research derived
from these models will not acknowledge limitations of this assumption,
or difficulty in the literature with quantifying an ecosystem scale
slope paramter \citep{Medlyn_2017}. Given that something so
fundamental as the ET resposne to VPD varies strongly with g$_1$, we
must invenst more in understanding g$_1$'s behavir at the ecosystem
scale, and the efficacy of models using g$_1$'s for ecosystem-scale
fluxes.

So far we have discussed the role of different parameters in the ET
response. However, the form of the ET response is imposed by our
choice of stomatal conductance model. We now examine how the model
choice, and speicifcally the exponent of the VPD dependence in teh
stomatal conductance model, can impact the general form of the ET-VPD
relationship. This is relevant as many models include an
exponententiated VPD term \citep[e.g.,][]{Leuning_1990}, but the value
of the exponent can vary. It is useful to understand how generally the
choice of model imposes the functional form of the ET-VPD
relationship, given its fundamental importance for land atmosphere
interaction.
\subsection{Functional form of ET dependence on VPD and its relation
  to the VPD exponent}
\label{functional_form}

>>>>

  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

In the ``sign term'' most of quantities are relatively constant,
except for VPD and the plant parameters g$_1$ and uWUE. In the scaling
term, most of the terms are relatively constant with the exception of
aerodynamic conductance ($g_a$) and temperature (especially its effect
through $\Delta$). To determine the range of probable ET responses to
VPD we will systematically vary this parameters according to Table
\ref{param_varying}, while all other parameters are held fixed (Table
\ref{param_fixed}). Physical variables ($g_a$, $T$) and plant
physiological parameters (g$_1$, uWUE) are varied according to
literature-based expectations for a range of growing season conditions
and plant types \citep{Zhou_2015, Medlyn_2017}. Using this previous
literature we can connect the effect of varying plant parameters to
specific plant types and characteristics.

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided in cases where the quantities are directly derived from
    previous literature. Conceptual percentiles
    for interpreting each quantity are noted.}
  \label{param_varying}
  \centering
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min (~15th percentile) & med (~50th percentile) & max
                                                                                            (~15th
                                                                                            percentile)
    & citation  \\
    \hline
    \input{param_varying.tex}
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}
  \centering \includegraphics{./fully_idealized_sign.pdf}
  \caption{The functional form of the $\frac{\partial \; ET}{\partial
      \; VPD}$ evaluated at the median value of the scaling term, for
    varying values of g$_1$ and uWUE as given in Table \ref{param_varying}.}
  \label{sign}
\end{figure}

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (through $\Delta$ controls the efficiency of
energy conversion to latent heat), amplify or suppress the plant
physiological response represented in the sign term. Both lower
temperatures and higher aerodynamic conductance lead to amplified ET
response to VPD, with the variability of aerodynamic conductance
resulting in a slightly higher variability of ET response relative to
temperature, which is still an important factor.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems. Our
results are a conceptual tool for answering the question, ``What is
the ET response to VPD with all other quantities held fixed?'' This is
useful, and we would argue is critical, for understanding complex
responses when many quantities are varying simultaneously. Through
partial derivatives, we can approach answering questions like ``What
is the relative contribution of VPD in some observed response given
that X Y and Z were also varying?'' In this example, X Y and Z could
be, for example, radiation, soil moisture, and temperature. Equation
\ref{d_et} explicitly provides an estimate of the ET response to VPD,
given assumptions about ecosystem plant characteristics through the
parameters g$_1$ and uWUE. Equation \ref{d_et} is generic to external
environmental factors and the timescale over which a VPD response
acts, and has a straightforward conceptual meaning. It provides the ET
response, given other quantities and parameters held fixed. So, while
the plant parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and that they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}) it is
useful to examine how these quantities vary in previous literature,
framed by our results on ET responses. In \citet{Zhou_2015} most plant
types have uWUE similar to our median value, where whether ET
response is positive (water intensive) or negative (water conservative)
depends on VPD. However, crops, which we expect to prioritize carbon
uptake over water conservation, have a higher uWUE closer to our
maximum value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which intuitively makes sense given
that we expect crops to keep stomata open for access to carbon, at the
cost of water loss. Shrubs, and to a lesser extent grass, gave an uWUE
closer to the minimum uWUE. For these plant types, we would then
expect a higher likelihood of a water conservative response; that is, a
decrease of ET in response to VPD. It is important to note that, while
\citet{Zhou_2015} did not examine the roll of soil moisture in within
plant-type variability of uWUE, we might expect some variation,
especially in extreme cases when soil water becomes a limiting factor.

In \citet{Medlyn_2017} there is not as robust of a relationship
between plant type and g$_1$ as there was for uWUE in
\citet{Zhou_2015}. Given that there is a robust relationship between
g$_1$ and plant type in \citet{Lin_2015}, the ambiguous results in
\citet{Medlyn_2017} could be due to noise in ecosystem scale
observations, the consequences of imposing a model structure in some
of the g$_1$ observations, or real ecosystem-scale within-plant type
variability. These differences between leaf-scale behavior and
ecosystem scale behavior highlight the importance of understanding how
and why slope parameters vary at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can make some
connections between g$_1$-induced ET response variability and plant
characteristics, which are subject to the strong caveat that we still
do not fully understand the g$_1$'s utility at the ecosystem
scale. Extrapolating \citet{Lin_2015}'s results (e.g. Fig. 2) to the
ecosystem scale would reveal the following relationships between g$_1$
and plant types:

\begin{itemize}
  \item C3 plants would have a generally higher g$_1$ than C4 plants,
  \item Crops would have a larger g$_1$ than shrubs, grass, and
    angiosperm trees, which have a higher g$_1$ than gymnosperm
    trees.
  \item Tropical and temperate climates would generally host plants
    with a higher g$_1$ than arctic and boreal climates.
\end{itemize}
For all of the above relationships, a higher g$_1$ means a higher
likelihood of a positive ET response to increasing VPD (water
intensive), and a smaller VPD dependence of the ET response.

Our theoretical results, when framed by previous literature tying
parameters to plant types and climate, highlight the variability of ET
response to VPD as a function of plant characteristics. Whether an
ecosystem increases or decreases ET in response to VPD will vary
depending on vegetation and climate. Generalizations about ET response
to VPD therefore require thoughtful consideration of ecosystem
characteristics; statements such as ``ET increases with warming due to
increases in VPD'' may be false depending on the water conservation
strategy of a given ecosystem **CITE SOME EXAMPLES OF THIS LOGIC FROM
GLOBAL WARMING LIT**.

Additionally, the sensitivity of ET response to plant parameter
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf and ecosystem scale
are not directly analogous much be understood. While the derivation
here was explicit about the assumptions of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. In particular, many models will assume a constant
g$_1$ term within a given plant functional type, and research derived
from these models will not acknowledge limitations of this assumption,
or difficulty in the literature with quantifying an ecosystem scale
slope parameter \citep{Medlyn_2017}. Given that something so
fundamental as the ET response to VPD varies strongly with g$_1$, we
must invest more in understanding g$_1$'s behavior at the ecosystem
scale, and the efficacy of models using g$_1$'s for ecosystem-scale
fluxes.

So far we have discussed the role of different parameters in the ET
response. However, the form of the ET response is imposed by our
choice of stomatal conductance model. We now examine how the model
choice, and specifically the exponent of the VPD dependence in the
stomatal conductance model, can impact the general form of the ET-VPD
relationship. This is relevant as many models include an
exponentiated VPD term \citep[e.g.,][]{Leuning_1990}, but the value
of the exponent can vary. It is useful to understand how generally the
choice of model imposes the functional form of the ET-VPD
relationship, given its fundamental importance for land atmosphere
interaction.
\subsection{Functional form of ET dependence on VPD and its relation
  to the VPD exponent}
\label{functional_form}

### 2019-02-17 doc/paper/conclusions.tex
spell check done and makefile updated
now proofread to send off to Pierre and Changjie

We derived a new form of Penman Monteith using the concept of
semi-empirical optimal stomatal regulation \citep{Lin_2015,
MEDLYN_2011} and near constant uWUE \citep{Zhou_2015} to remove the
implicit dependence of stomatal conductance on GPP and ET. With our
new form of Penman Monteith we developed a theory for when an
ecosystem will tend to reduce or increase ET with increasing VPD,
which we framed using previous literature exploring the relatiohsip
between plant parameters, plant types, and climate. The goal was to
understand the range of likely ET responses to VPD and develop some
intuition for how the ET response may vary with plant types and
climate. This intuition can be used to disentangle land atmosphere
feedbacks in more complicated scenarios, and will aid interpretation
of observations and more sophisticated models.

ET response to VPD can vary from strongly water conservative (ET
always decreasing in response to increasing VPD), to strongly water
intensive (ET always increasing in response to VPD), which is
idicative of the large observed diversity of stomatal regulation and
plany physiological strategy. Higher uWUE and g$_1$ values make the ET
response more likely to be positive, while decreasing temperature and
increasing g$_a$ amplify the magnitue of the response. Previous
literature \citep{Zhou_2015, Lin_2015} suggests that crops, through
association with higher g$_1$ and uWUE, are more likely to exhibit a
postiive ET response to VPD. Shrubs (lower uWUE), C4 plants (lower
g$_1$), gymnosperm trees (lower g$_1$), and plants in a artic and
boreal climates (lower g$_1$ are more likely to exhibit a negative ET
response to VPD. However interpretation of g$_1$-induced variability
is partially muddied by ambiguity in upscaling leaf-scale estimations
of g$_1$ to the ecosystem scale \citep{Medlyn_2017}.

Our paper builds important intuition for how plants respond to VPD
perturbations. We show that given optimal stomatal function and fixed
environmental conditions, the ET-VPD dependence is theoretically
concave upward, with ET increasing with increasing VPD as VPD
increases past some critical value **Add VPD_crit plot to
results??**. However future research should focus on fully
understanding the form of stomatal VPD dependence, as this result is
sensitive to the exponent of VPD dependence, which we currently
believe is 1/2 \citep{MEDLYN_2011, Zhou_2014}. Indeed, this
sensitivity to the exponent of VPD dependence is an important result
itself: land surface models, including those used in earth system
models for climate forecasts, employ different assumptions about the
exponent of VPD dependence in stomatal conductance
\citep[e.g.,][]{Ball_1987, Leuning_1990, MEDLYN_2011}, and these
assumptions can fundamentally change the relationship between ET and
VPD from one that is concave upward (local minimum in ET) to one that
is concave downward (local maximum in ET).

Our results are also applicable to understanding the impact of
expected increases in VPD induced by global change.  Plant
physiological responses to direct CO$_2$ effects
\citep[e.g.,][]{Swann_2016, Lemordant_2018} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide a framework for understanding ET response to ET
using a simplfied model fo two plant parameters. Feedbacks between the
land and the atmosphere may alter the net response to a long-timescale
global VPD perturbation, but our focus on the one way plant response
to a VPD perturbation in the atmospheric boundary layer is an
important first step to disentangling such feedbacks, both in
observations and model simulations of the present and future. By
removing Penman Monteith's dependence on implicit relationships
between GPP, VPD, and ET, we allow for explicit, accurate future
analysis of plant-VPD feedbacks in the atmospheric boundary layer
(Eq. (\ref{full})). Our approach can be extended to examine varying
plant response to more nuanced consideration of plant type and
climate. Any plant physiological heterogeneity or feedback that can be
conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015,
Medlyn_2017} and/or uWUE \citep[e.g.][]{Zhou_2014} are representable
within our framework, which opens the door for a hierarchy of more
sophisticated climate- and plant-specific analysis of ET sensitivity
to environmental variables (including VPD). We argue that such
simplified conceptual frameworks are critical tools for disentangling
land-atmosphere feedbacks at various scales, from diurnal to seasonal
and beyond, and to characterize ET response in a warmer,
atmospherically drier, and enriched CO$_2$ world.

>>>>

We derived a new form of Penman Monteith using the concept of
semi-empirical optimal stomatal regulation \citep{Lin_2015,
MEDLYN_2011} and near constant uWUE \citep{Zhou_2015} to remove the
implicit dependence of stomatal conductance on GPP and ET. With our
new form of Penman Monteith we developed a theory for when an
ecosystem will tend to reduce or increase ET with increasing VPD,
which we framed using previous literature exploring the relationship
between plant parameters, plant types, and climate. The goal was to
understand the range of likely ET responses to VPD and develop some
intuition for how the ET response may vary with plant types and
climate. This intuition can be used to disentangle land atmosphere
feedbacks in more complicated scenarios, and will aid interpretation
of observations and more sophisticated models.

ET response to VPD can vary from strongly water conservative (ET
always decreasing in response to increasing VPD), to strongly water
intensive (ET always increasing in response to VPD), which is
indicative of the large observed diversity of stomatal regulation and
plant physiological strategy. Higher uWUE and g$_1$ values make the ET
response more likely to be positive, while decreasing temperature and
increasing g$_a$ amplify the magnitude of the response. Previous
literature \citep{Zhou_2015, Lin_2015} suggests that crops, through
association with higher g$_1$ and uWUE, are more likely to exhibit a
positive ET response to VPD. Shrubs (lower uWUE), C4 plants (lower
g$_1$), gymnosperm trees (lower g$_1$), and plants in a arctic and
boreal climates (lower g$_1$ are more likely to exhibit a negative ET
response to VPD. However interpretation of g$_1$-induced variability
is partially muddied by ambiguity in upscaling leaf-scale estimations
of g$_1$ to the ecosystem scale \citep{Medlyn_2017}.

Our paper builds important intuition for how plants respond to VPD
perturbations. We show that given optimal stomatal function and fixed
environmental conditions, the ET-VPD dependence is theoretically
concave upward, with ET increasing with increasing VPD as VPD
increases past some critical value **Add VPD crit plot to
results??**. However future research should focus on fully
understanding the form of stomatal VPD dependence, as this result is
sensitive to the exponent of VPD dependence, which we currently
believe is 1/2 \citep{MEDLYN_2011, Zhou_2014}. Indeed, this
sensitivity to the exponent of VPD dependence is an important result
itself: land surface models, including those used in earth system
models for climate forecasts, employ different assumptions about the
exponent of VPD dependence in stomatal conductance
\citep[e.g.,][]{Ball_1987, Leuning_1990, MEDLYN_2011}, and these
assumptions can fundamentally change the relationship between ET and
VPD from one that is concave upward (local minimum in ET) to one that
is concave downward (local maximum in ET).

Our results are also applicable to understanding the impact of
expected increases in VPD induced by global change.  Plant
physiological responses to direct CO$_2$ effects
\citep[e.g.,][]{Swann_2016, Lemordant_2018} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide a framework for understanding ET response to ET
using a simplified model of two plant parameters. Feedbacks between the
land and the atmosphere may alter the net response to a long-timescale
global VPD perturbation, but our focus on the one way plant response
to a VPD perturbation in the atmospheric boundary layer is an
important first step to disentangling such feedbacks, both in
observations and model simulations of the present and future. By
removing Penman Monteith's dependence on implicit relationships
between GPP, VPD, and ET, we allow for explicit, accurate future
analysis of plant-VPD feedbacks in the atmospheric boundary layer
(Eq. (\ref{full})). Our approach can be extended to examine varying
plant response to more nuanced consideration of plant type and
climate. Any plant physiological heterogeneity or feedback that can be
conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015,
Medlyn_2017} and/or uWUE \citep[e.g.][]{Zhou_2014} are representable
within our framework, which opens the door for a hierarchy of more
sophisticated climate- and plant-specific analysis of ET sensitivity
to environmental variables (including VPD). We argue that such
simplified conceptual frameworks are critical tools for disentangling
land-atmosphere feedbacks at various scales, from diurnal to seasonal
and beyond, and to characterize ET response in a warmer,
atmospherically drier, and enriched CO$_2$ world.
